---
title: "*Spatial Statistics Lab 9*"
author: '**Onyedikachi J Okeke**'
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **0.0 To Load the library**
```{r,comment="",echo=TRUE,eval=FALSE}
library(sp)
library(spatstat)
library(sf)
library(spatstat.geom)
library(ctv)
library(terra)
library(spdep)
library(rgdal)
library(rgdal)
library(terra)
library(RColorBrewer)
library(classInt)
library(epitools)
library(DCluster)
library(lmtest)
library(car)
library(spatialreg)
library(spdep)
library(classInt)
library(spgwr)
library(ggplot2)
library(gstat)
library(lattice)
```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}

library(sp)
library(spatstat)
library(sf)
library(spatstat.geom)
library(ctv)
library(terra)
library(spdep)
library(rgdal)
library(rgdal)
library(terra)
library(RColorBrewer)
library(classInt)
library(epitools)
library(DCluster)
library(lmtest)
library(car)
library(spatialreg)
library(spdep)
library(classInt)
library(spgwr)
library(ggplot2)
library(gstat)
library(lattice)

```

### **0.1 Loading data set comprises of four heavy metals (e.g. zinc) measured in the top soil.**
```{r,comment="",echo=TRUE,eval=FALSE}
library(lattice)
library(sp)
data(meuse) #use the system file
coordinates(meuse) <- c("x", "y")
#map the zinc value
data(meuse.grid)
spplot(meuse, "zinc", do.log = T)
#create a raster surface of distance to river
coordinates(meuse.grid) = ~x+y
proj4string(meuse.grid) <- CRS("+init=epsg:28992")
gridded(meuse.grid) = TRUE
spplot(meuse.grid)
image(meuse.grid["dist"])
title("distance to river (red = 0)")

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}

library(lattice)
library(sp)
data(meuse) #use the system file
coordinates(meuse) <- c("x", "y")
#map the zinc value
data(meuse.grid)
spplot(meuse, "zinc", do.log = T)
#create a raster surface of distance to river
coordinates(meuse.grid) = ~x+y
proj4string(meuse.grid) <- CRS("+init=epsg:28992")
gridded(meuse.grid) = TRUE
spplot(meuse.grid)
image(meuse.grid["dist"])
title("distance to river (red = 0)")

```
Q1. Take a screen shot of the two plots. Do you see any potential correlation between the zinc concentration and distance to river? Explain if any.

## **Fit a linear regression to model the relationship between zinc and distance to the river. Plot maps with fitted values and with residuals.**
```{r,comment="",echo=TRUE,eval=FALSE}

print(xyplot(log(zinc)~sqrt(dist), as.data.frame(meuse), asp = 1), split = c(1,1,2,1), more = TRUE)
zn.lm <- lm(log(zinc)~sqrt(dist), meuse)
meuse$fitted.s <- predict(zn.lm, meuse) - mean(predict(zn.lm, meuse))
meuse$residuals <- residuals(zn.lm)
print(spplot(meuse, c("fitted.s", "residuals"), cuts=4, col.regions=grey.colors(5, 0.95, 0.55, 2.2)), split = c(2,1,2,1))
```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}

print(xyplot(log(zinc)~sqrt(dist), as.data.frame(meuse), asp = 1), split = c(1,1,2,1), more = TRUE)
zn.lm <- lm(log(zinc)~sqrt(dist), meuse)
meuse$fitted.s <- predict(zn.lm, meuse) - mean(predict(zn.lm, meuse))
meuse$residuals <- residuals(zn.lm)
print(spplot(meuse, c("fitted.s", "residuals"), cuts=4, col.regions=grey.colors(5, 0.95, 0.55, 2.2)), split = c(2,1,2,1))

```

Q2. Take a screen shot of the plots. Describe the pattern of the residuals

## **converting into a SpatialPixels DataFrame**
```{r,comment="",echo=TRUE,eval=FALSE}
library(gstat)
data(meuse.grid)
coordinates(meuse.grid) <- c("x", "y")
meuse.grid <- as(meuse.grid, "SpatialPixelsDataFrame")
###inverse distance weighted interpolation

idw.out <- idw(zinc~1, meuse, meuse.grid, idp = 2.5)
as.data.frame(idw.out)[1:5,]
spplot(idw.out["var1.pred"], main = "zinc inverse distance weighted interpolations")

idw.out2 <- idw(zinc~1, meuse, meuse.grid, idp = 5)
as.data.frame(idw.out2)[1:5,]
spplot(idw.out2["var1.pred"], main = "zinc inverse distance weighted interpolations, idp=5")

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}
library(gstat)
data(meuse.grid)
coordinates(meuse.grid) <- c("x", "y")
meuse.grid <- as(meuse.grid, "SpatialPixelsDataFrame")
###inverse distance weighted interpolation

idw.out <- idw(zinc~1, meuse, meuse.grid, idp = 2.5)
as.data.frame(idw.out)[1:5,]
spplot(idw.out["var1.pred"], main = "zinc inverse distance weighted interpolations")


idw.out2 <- idw(zinc~1, meuse, meuse.grid, idp = 5)
as.data.frame(idw.out2)[1:5,]
spplot(idw.out2["var1.pred"], main = "Zinc inverse distance weighted interpolations, idp=5")

```

Q3. Take a screen shot of the plot. Modify the idp value to 5 and rerun the above statements. Compare the result against the first one.

## **Variograms calculations**
```{r,comment="",echo=TRUE,eval=FALSE}
library(lattice)

log(zinc)~1 
#means that we assume a constant trend for the variable log(zinc).
svgm <- variogram(log(zinc) ~ 1, meuse)

svgm

cld <- variogram(log(zinc) ~ 1, meuse, cloud = TRUE)
svgm <- variogram(log(zinc) ~ 1, meuse)
d <- data.frame(gamma = c(cld$gamma, svgm$gamma),
dist = c(cld$dist, svgm$dist),
id = c(rep("cloud", nrow(cld)), rep("sample variogram", nrow(svgm)))
)
xyplot(gamma ~ dist | id, d,
scales = list(y = list(relation = "free", ylim = list(NULL, c(-.005,0.7)))),
layout = c(1, 2), as.table = TRUE,
panel = function(x,y, ...) {
if (panel.number() == 2)
ltext(x+10, y, svgm$np, adj = c(0,0.5)) #$
panel.xyplot(x,y,...)
},
xlim = c(0, 1590),
cex = .5, pch = 3
)

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}

library(lattice)

log(zinc)~1 
#means that we assume a constant trend for the variable log(zinc).
svgm <- variogram(log(zinc) ~ 1, meuse)

svgm

cld <- variogram(log(zinc) ~ 1, meuse, cloud = TRUE)
svgm <- variogram(log(zinc) ~ 1, meuse)
d <- data.frame(gamma = c(cld$gamma, svgm$gamma),
dist = c(cld$dist, svgm$dist),
id = c(rep("cloud", nrow(cld)), rep("sample variogram", nrow(svgm)))
)
xyplot(gamma ~ dist | id, d,
scales = list(y = list(relation = "free", ylim = list(NULL, c(-.005,0.7)))),
layout = c(1, 2), as.table = TRUE,
panel = function(x,y, ...) {
if (panel.number() == 2)
ltext(x+10, y, svgm$np, adj = c(0,0.5)) #$
panel.xyplot(x,y,...)
},
xlim = c(0, 1590),
cex = .5, pch = 3
)

```

Q4. Take a screen shot of the plot. What are the differences between a variogram cloud (top one) and empirical variogram (button one).


## **Cutoff, Lag Width, Direction Dependence**
```{r,comment="",echo=TRUE,eval=FALSE}

plot(variogram(log(zinc) ~ 1, meuse))

plot(variogram(log(zinc) ~ 1, meuse, alpha = c(0, 45, 90, 135)))

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}

plot(variogram(log(zinc) ~ 1, meuse))

plot(variogram(log(zinc) ~ 1, meuse, alpha = c(0, 45, 90, 135)))

```

Q5. Take a screen shot of the plot. Describe and compare the patterns in the four subplots.

## **Variogram Modelling**
```{r,comment="",echo=TRUE,eval=FALSE}
v <- variogram(log(zinc) ~ 1, meuse)
plot(v)
v <- variogram(log(zinc) ~ 1, meuse)
v.fit <- fit.variogram(v, vgm(1, "Sph", 800, 1)) ##spherical model
plot(v, v.fit, pch = 3)

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}
v <- variogram(log(zinc) ~ 1, meuse)
plot(v)
v <- variogram(log(zinc) ~ 1, meuse)
v.fit <- fit.variogram(v, vgm(1, "Sph", 800, 1)) ##spherical model
plot(v, v.fit, pch = 3)

```

Q6. Take a screen shot of the plot.

## **Classify directions into four direction intervals**
```{r,comment="",echo=TRUE,eval=FALSE}

v.dir <- variogram(log(zinc)~1,meuse,alpha=(0:3)*45)
v.anis <- vgm(.6, "Sph", 1600, .05, anis=c(45, 0.3))
print(plot(v.dir, v.anis, pch=3))

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}

v.dir <- variogram(log(zinc)~1,meuse,alpha=(0:3)*45)
v.anis <- vgm(.6, "Sph", 1600, .05, anis=c(45, 0.3))
print(plot(v.dir, v.anis, pch=3))

```

Q7. Take a screen shot of the plot.

## **Simple, Ordinary, and Universal Kriging**
```{r,comment="",echo=TRUE,eval=FALSE}

lz.sk <- krige(log(zinc)~1, meuse, meuse.grid, v.fit, beta = 5.9)
lz.ok <- krige(log(zinc)~1, meuse, meuse.grid, v.fit)
lz.uk <- krige(log(zinc)~sqrt(dist), meuse, meuse.grid, v.fit)
plot(lz.sk)
plot(lz.ok)
plot(lz.uk)

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}

lz.sk <- krige(log(zinc)~1, meuse, meuse.grid, v.fit, beta = 5.9)
lz.ok <- krige(log(zinc)~1, meuse, meuse.grid, v.fit)
lz.uk <- krige(log(zinc)~sqrt(dist), meuse, meuse.grid, v.fit)
plot(lz.sk)
plot(lz.ok)
plot(lz.uk)

```

Q8. Take a screen shot of these plots and describe their differences.

## **Model Diagnostics (cross validation)**
```{r,comment="",echo=TRUE,eval=FALSE}

sel100 <- sample(1:155, 100)
m.model <- meuse[sel100,]
m.valid <- meuse[-sel100,]
v100.fit <- fit.variogram(variogram(log(zinc)~1, m.model), vgm(1, "Sph", 800, 1))
m.valid.pr <- krige(log(zinc)~1, m.model, m.valid, v100.fit) #[using ordinary kriging]
resid.kr <- log(m.valid$zinc) - m.valid.pr$var1.pred
summary(resid.kr)
resid.mean <- log(m.valid$zinc) - mean(log(m.valid$zinc))
R2 <- 1 - sum(resid.kr^2)/sum(resid.mean^2)
R2

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}

sel100 <- sample(1:155, 100)
m.model <- meuse[sel100,]
m.valid <- meuse[-sel100,]
v100.fit <- fit.variogram(variogram(log(zinc)~1, m.model), vgm(1, "Sph", 800, 1))
m.valid.pr <- krige(log(zinc)~1, m.model, m.valid, v100.fit) #[using ordinary kriging]
resid.kr <- log(m.valid$zinc) - m.valid.pr$var1.pred
summary(resid.kr)
resid.mean <- log(m.valid$zinc) - mean(log(m.valid$zinc))
R2 <- 1 - sum(resid.kr^2)/sum(resid.mean^2)
R2

```